---
title: "SN_ASR_DRAFT_analysis — MIRRORED the LC for the SN (includes future rostral/caudal & left/right stubs)"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    output_file: 'SN_ASR_DRAFT_analysis_`r format(Sys.time(), "%Y%m%d")`.html'
output_dir: MRI_data/analysis
---

```{r global_options, include=FALSE}
# Clears the R environment so every knit starts from the same clean state.
rm(list=ls()) # clear global environment (as in original)
library(knitr)
#knitr::opts_chunk$set(fig.width=3, fig.height=3, warning=FALSE, message=FALSE)
#opts_knit$set(global.par = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(file.path(dirname(knitr::current_input()), '..')))
options(knitr.duplicate.label = 'allow')
```

```{r include=FALSE}
# This keeps the margin tweak hidden but ensures later plots have the right spacing.
par(mar=c(5,5,0,0)) # keep separate, as in original
```

```{r libraries, message=FALSE, warning=FALSE}
# This loads all plotting and mixed-model packages needed for the mirrored SN analyses.
# Packages used in the original LC analyses + mirrors here
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(mgcv)
library(gratia)
library(broom.mixed)
library(scales)
library(janitor)
library(reshape2)
library(ggokabeito)  # scale_color_okabe_ito()
library(sjPlot)      # tab_model(), plot_model()
```

## Plotting parameters and theme (unchanged from original)
```{r}
# This defines the shared plot formatting so every figure follows the same style.
text.size = 10
se.width = .075

ggplot.theme <- theme(
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank(),
  panel.background = element_blank(),
  panel.spacing = unit(0.2,'in'),
  axis.line.x = element_line(size = 0.5, colour = "black"),
  axis.line.y = element_line(size = 0.5, colour = "black"),
  axis.title = element_text(size = text.size),
  axis.text = element_text(size = text.size),
  strip.background = element_rect(colour = 'black'),
  strip.text = element_text(size = text.size),
  legend.direction='horizontal',
  legend.spacing=unit(.02,'in'),
  legend.position = 'top',
  legend.background = element_blank(),
  legend.margin = margin(rep(.1,4)),
  legend.key = element_rect(fill=NA),
  legend.title = element_blank(),
  legend.text = element_text(size = 12),
  legend.text.align = 1
)
```

## Load datasets (SN = target; cerebellum kept as control)

This section loads the neuromelanin summary tables. BLACK ANT writes the SN metrics
to `${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/`, so the script
prefers that path but falls back to the legacy MOTIP locations when necessary. The
same fallback logic applies to the cerebellum control table so collaborators on
different machines can knit without editing paths.
```{r load-data}
# This pulls the SN and cerebellum measurements off disk so the analysis has data to work with.
# RELIES ON: literal file paths (but can be overridden via SN_ASR_SN_TABLE/SN_ASR_SN_LABEL env vars)
sn_override <- Sys.getenv("SN_ASR_SN_TABLE")
sn_label_override <- Sys.getenv("SN_ASR_SN_LABEL")

resolve_repo_path <- function(path) {
  if (!nzchar(path)) {
    return("")
  }
  if (grepl("^/", path)) {
    return(path)
  }
  base_dir <- tryCatch({
    if ("knitr" %in% loadedNamespaces()) {
      input_file <- knitr::current_input()
      if (nzchar(input_file)) {
        return(dirname(input_file))
      }
    }
    getwd()
  }, error = function(e) getwd())
  normalizePath(file.path(base_dir, "..", path), winslash = "/", mustWork = FALSE)
}

expand_candidates <- function(paths) {
  expanded <- character()
  for (pattern in paths) {
    if (grepl("\\*", pattern)) {
      matches <- Sys.glob(pattern)
      if (length(matches) > 1) {
        info <- file.info(matches)
        matches <- matches[order(info$mtime, decreasing = TRUE)]
      }
      expanded <- c(expanded, matches)
    } else {
      expanded <- c(expanded, pattern)
    }
  }
  unique(expanded)
}

sn_candidate_patterns <- c("${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_V4_*.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_V4_latest.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_V3_*.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_V3_latest.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_V3.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved_2.txt",
                           "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/SN_nmdata_autosaved.txt",
                           "${MOTIP_ROOT:-/path/to/MOTIP2018}/scripts/tse/SN_nmdata_autosaved.txt",
                           "${MOTIP_ROOT:-/path/to/MOTIP2018}/scripts/tse/SN_nmdata_autosaved.txt")

sn_candidates <- expand_candidates(sn_candidate_patterns)

sn_path <- if (nzchar(sn_override)) {
  resolve_repo_path(sn_override)
} else {
  existing <- sn_candidates[file.exists(sn_candidates)]
  if (length(existing) == 0) "" else existing[1]
}

if (!nzchar(sn_path) || !file.exists(sn_path)) {
  stop("Could not locate SN_nmdata_autosaved*.txt in SNHEART or MOTIP folders")
}

sn_dataset_label <- if (nzchar(sn_label_override)) {
  sn_label_override
} else if (grepl("V4", basename(sn_path))) {
  "BlackANT_V4"
} else if (grepl("V3", basename(sn_path))) {
  "BlackANT_V3"
} else if (grepl("_2", basename(sn_path))) {
  "BlackANT_V2"
} else {
  "BlackANT_V1"
}
sn_dataset_slug <- gsub("[^A-Za-z0-9]+", "", sn_dataset_label)
if (!nzchar(sn_dataset_slug)) sn_dataset_slug <- "SN"
sn_dataset_path <- sn_path
message(sprintf("Using SN dataset: %s (%s)", sn_dataset_label, sn_dataset_path))

cb_primary   <- "${SNHEART_ROOT:-/path/to/SNHEART}/MRI_data/analysis/cerebellumdata_autosaved.txt"
cb_fallbacks <- c("${MOTIP_ROOT:-/path/to/MOTIP2018}/scripts/tse/cerebellumdata_autosaved.txt",
                  "${MOTIP_ROOT:-/path/to/MOTIP2018}/scripts/tse/cerebellumdata_autosaved.txt")
cb_path <- if (file.exists(cb_primary)) {
  cb_primary
} else {
  found <- cb_fallbacks[file.exists(cb_fallbacks)]
  if (length(found) == 0) "" else found[1]
}
if (!nzchar(cb_path)) {
  stop("Could not locate cerebellumdata_autosaved.txt in SNHEART or MOTIP folders")
}

read_csv_file <- function(path) {
  if (requireNamespace("readr", quietly = TRUE)) {
    readr::read_csv(path, show_col_types = FALSE)
  } else {
    read.table(path, sep = ",", header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
  }
}

SN <- as.data.frame(read_csv_file(sn_path))
cerebellum <- as.data.frame(read_csv_file(cb_path))
if (!"subject" %in% names(SN)) {
  SN$subject <- if ("sub" %in% names(SN)) SN[["sub"]] else NA
}
if (!"subject" %in% names(cerebellum)) {
  cerebellum$subject <- if ("sub" %in% names(cerebellum)) cerebellum[["sub"]] else NA
}

```

> **SN dataset in use:** `r sn_dataset_label` (`r sn_dataset_path`)

```{r echo=FALSE, results='asis'}
if (identical(sn_dataset_label, "BlackANT_V3")) {
  cat("<em>V3 note:</em> CNR still uses the raw TSE signal with the original control ROI, while the neuromelanin intensity column now averages the corrected image over only the top 30% of probability-weighted SN voxels. That masking difference can make intensity and CNR move in opposite directions and does not indicate a calculation error.\n")
} else if (identical(sn_dataset_label, "BlackANT_V4")) {
  cat("<em>V4 note:</em> SN and control masks now keep the same percentile and the exported CNR column is computed on the corrected neuromelanin image (matching the intensity column). Use `CNR_raw` if you need the legacy raw-grid calculation.\n")
}
```

colnames(SN)[1] <- "subject"
colnames(cerebellum)[1] <- "subject"

SN <- SN %>% mutate(subject = as.numeric(subject))

```{r ensure-age, message=FALSE}
# Ensure age exists before the low-signal summaries. The uber roster holds the
# demographic columns used later in the report, but some SN tables (including
# the new Irredeemable exports) arrive without age baked in. Pull just the age
# column now so the early chunks can sort/label the low group before the full
# covariate merge runs inside the section body.
if (!"age" %in% names(SN)) {
  uber_paths <- c("${SNHEART_ROOT:-/path/to/SNHEART}/behavioral/uber_data_20240704.csv",
                  "${MOTIP_ROOT:-/path/to/MOTIP2018}/uber_data_20240704.csv",
                  "${MOTIP_ROOT:-/path/to/MOTIP2018}/uber_data_20240704.csv",
                  "${MOTIP_ROOT:-/path/to/MOTIP2018}/uber_data_20240704.csv")
  uber_file <- uber_paths[file.exists(uber_paths)][1]
  if (!is.na(uber_file)) {
    uber_age <- as.data.frame(read_csv_file(uber_file)) %>%
      dplyr::select(subject, age)
    SN <- SN %>%
      left_join(uber_age, by = "subject")
  } else {
    SN$age <- NA_real_
  }
}
```

```{r low30-flags, message=FALSE}
calc_residuals <- function(df, response) {
  required <- c(response, "age", "gender", "race")
  if (!all(required[-1] %in% names(df))) {
    return(rep(NA_real_, nrow(df)))
  }
  complete_idx <- complete.cases(df[, required])
  residuals <- rep(NA_real_, nrow(df))
  if (sum(complete_idx) < 12) {
    return(residuals)
  }
  formula <- as.formula(paste(response, "~ age + gender + race"))
  fit <- try(lm(formula, data = df[complete_idx, ]), silent = TRUE)
  if (inherits(fit, "try-error")) {
    return(residuals)
  }
  residuals[complete_idx] <- resid(fit)
  residuals
}

if (!"gender" %in% names(SN)) {
  if ("Gender" %in% names(SN)) {
    SN$gender <- SN$Gender
  } else {
    SN$gender <- factor(rep(NA_character_, nrow(SN)))
  }
}
if (!"race" %in% names(SN)) {
  SN$race <- factor(rep(NA_character_, nrow(SN)))
}
SN$gender <- droplevels(factor(SN$gender))
SN$race <- droplevels(factor(SN$race))

low_overall_pct <- 0.30
low_within_pct <- 0.20
age_breaks <- c(18, 30, 50, 60, 70, Inf)
age_labels <- c("18-30", "30-50", "50-60", "60-70", "70+")

SN <- SN %>%
  mutate(
    CNR_residual = calc_residuals(., "CNR"),
    intensity_residual = calc_residuals(., "intensity")
  )

cnr_resid_cut <- quantile(SN$CNR_residual, low_overall_pct, na.rm = TRUE)
int_resid_cut <- quantile(SN$intensity_residual, low_overall_pct, na.rm = TRUE)

SN <- SN %>%
  mutate(
    low_cnr_residual = if (is.finite(cnr_resid_cut)) CNR_residual <= cnr_resid_cut else NA,
    low_intensity_residual = if (is.finite(int_resid_cut)) intensity_residual <= int_resid_cut else NA
  )

if ("age" %in% names(SN)) {
  SN$age_decile <- cut(
    SN$age,
    breaks = age_breaks,
    include.lowest = TRUE,
    right = FALSE,
    labels = age_labels
  )
}

SN <- SN %>%
  group_by(gender, race, age_decile) %>%
  mutate(
    low_cnr_within = percent_rank(CNR) <= low_within_pct,
    low_intensity_within = percent_rank(intensity) <= low_within_pct
  ) %>%
  ungroup()

SN <- SN %>%
  mutate(
    low_cnr = coalesce(low_cnr_within, low_cnr_residual, FALSE),
    low_intensity = coalesce(low_intensity_within, low_intensity_residual, FALSE),
    low_both = low_cnr & low_intensity,
    low15_flag = low_cnr | low_intensity
  )

low_overlap <- SN %>%
  count(low_cnr, low_intensity) %>%
  arrange(desc(n))

percent_overlap <- SN %>%
  summarize(pct = mean(low_both, na.rm = TRUE)) %>%
  pull(pct)
```

The residual-based 30th-percentile cutoffs land at `r signif(cnr_resid_cut, 3)` for CNR and `r signif(int_resid_cut, 3)` for intensity. Using the within-group 20th-percentile screen plus the residual threshold, `r scales::percent(mean(SN$low_cnr, na.rm = TRUE), accuracy = 0.1)` of the cohort is flagged as low CNR, `r scales::percent(mean(SN$low_intensity, na.rm = TRUE), accuracy = 0.1)` as low intensity, and `r scales::percent(percent_overlap, accuracy = 0.1)` fall into both groups.

```{r low30-table, echo=FALSE}
low_overlap
```

```{r low30-subjects, echo=FALSE}
SN %>%
  filter(low_both) %>%
  arrange(age) %>%
  select(subject, age, CNR, intensity) %>%
  head(15)
```

These tables call out the overlapping low-signal subjects so we can sanity-check whether the same people drive both low-CNR and low-intensity cohorts.


# Section A — Thresholded SN metrics

Section A reruns the full SN analyses using the thresholded (> gcount cutoff) masks from Irredeemable Black ANT. All plots/tables below match the historical reports but now live under a dedicated heading.

```{r sectionA-setup, include=FALSE}
SN_variant_base <- SN
cerebellum_variant_base <- cerebellum
prepare_sn_variant <- function(sn_df, variant = c('thresholded','unthresholded')) {
  variant <- match.arg(variant)
  renamed <- sn_df
  if (variant == 'unthresholded') {
    rename_map <- c(
      CNR = 'CNR_unthresholded',
      CNR_peduncle = 'CNR_peduncle_unthresholded',
      total_intensity = 'total_intensity_unthresholded',
      intensity = 'intensity_unthresholded',
      background_intensity = 'background_intensity_unthresholded',
      total_volume = 'total_volume_unthresholded',
      intensity_minimum = 'intensity_minimum_unthresholded',
      intensity_maximum = 'intensity_maximum_unthresholded',
      sn_voxels = 'sn_voxels_unthresholded',
      sn_prob_mean = 'sn_prob_mean_unthresholded',
      mean_raw_sn = 'mean_raw_sn_unthresholded',
      mean_raw_control = 'mean_raw_control_unthresholded',
      std_raw_control = 'std_raw_control_unthresholded',
      mean_raw_peduncle = 'mean_raw_peduncle_unthresholded',
      bright_voxels = 'bright_voxels_unthresholded',
      bright_fraction = 'bright_fraction_unthresholded'
    )
    for (base_name in names(rename_map)) {
      alt_name <- rename_map[[base_name]]
      if (alt_name %in% names(renamed)) {
        renamed[[base_name]] <- renamed[[alt_name]]
      }
    }
  }
  renamed
}
SN <- prepare_sn_variant(SN_variant_base, 'thresholded')
cerebellum <- cerebellum_variant_base
sn_variant_label <- paste(sn_dataset_label, '(thresholded)')
SN <- SN %>%
  mutate(age_decile = cut(
    age,
    breaks = age_breaks,
    include.lowest = TRUE,
    right = FALSE,
    labels = age_labels
  ))
```

```{r child='Scripts/SN_ASR_section_body.Rmd'}
```

# Section B — Unthresholded SN metrics

Section B repeats every test after swapping in the unthresholded probabilistic masks (any voxel with >0 votes). Use these outputs when comparing against legacy autosaves or when you need the full SN volume without the gcount cutoff.

```{r sectionB-setup, include=FALSE}
SN <- prepare_sn_variant(SN_variant_base, 'unthresholded')
cerebellum <- cerebellum_variant_base
sn_variant_label <- paste(sn_dataset_label, '(unthresholded)')
SN <- SN %>%
  mutate(age_decile = cut(
    age,
    breaks = age_breaks,
    include.lowest = TRUE,
    right = FALSE,
    labels = age_labels
  ))
```

```{r child='Scripts/SN_ASR_section_body.Rmd'}
```
