---
title: "THUNDERBOLTS V5 CP-Only"
subtitle: "SNHEART Irredeemable CP-mask analyses without nigrosomes"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: null
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

# Section 01 — Plotting Parameters and Theme

```{r plotting-theme}
library(tidyverse)
library(glue)
library(scales)
library(patchwork)

text.size <- 10
gg_theme_sn_asr <- theme(
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank(),
  panel.background = element_blank(),
  panel.spacing = unit(0.3, 'in'),
  axis.line.x = element_line(size = 0.5, colour = 'black'),
  axis.line.y = element_line(size = 0.5, colour = 'black'),
  axis.title = element_text(size = text.size),
  axis.text = element_text(size = text.size),
  strip.background = element_rect(colour = 'black'),
  strip.text = element_text(size = text.size),
  legend.direction = 'horizontal',
  legend.spacing = unit(0.02, 'in'),
  legend.position = 'top',
  legend.background = element_blank(),
  legend.margin = margin(rep(0.1, 4)),
  legend.key = element_rect(fill = NA),
  legend.title = element_blank(),
  legend.text = element_text(size = 12),
  legend.text.align = 1,
  plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
)

wrap_text <- function(text, width = 90) stringr::str_wrap(text, width = width)

describe_test <- function(title, question, details) {
  cat(glue("**Test: {title}.** {wrap_text(question)}\n\n_Key setup_: {wrap_text(details)}\n\n"))
}

describe_plot <- function(title, interpretation) {
  cat(glue("**How to read the plot:** {wrap_text(title)}. {wrap_text(interpretation)}\n\n"))
}
```

# Section 02 — Helper Functions

```{r helper-functions}
read_delim_auto <- function(path) {
  if (!file.exists(path)) {
    stop(glue("File not found: {path}"))
  }
  if (grepl("\\.tsv$", path)) {
    readr::read_tsv(path, show_col_types = FALSE)
  } else {
    readr::read_csv(path, show_col_types = FALSE)
  }
}

first_existing <- function(paths) {
  matches <- paths[file.exists(paths)]
  if (length(matches) == 0) return(NA_character_)
  matches[[1]]
}

clean_factor <- function(values, drop_levels = c('NaN', 'NA', 'Unknown', '')) {
  factor_values <- factor(values)
  factor_values[factor_values %in% drop_levels] <- NA
  forcats::fct_drop(factor_values)
}

resolve_numeric_column <- function(df, candidates) {
  if (nrow(df) == 0) return(numeric())
  resolved <- rep(NA_real_, nrow(df))
  for (nm in candidates) {
    if (nm %in% names(df)) {
      vals <- suppressWarnings(as.numeric(df[[nm]]))
      replace_idx <- is.na(resolved)
      resolved[replace_idx] <- vals[replace_idx]
    }
  }
  resolved
}

resolve_gender_column <- function(df) {
  if (nrow(df) == 0) return(character())
  resolved <- rep(NA_character_, nrow(df))
  candidates <- c('gender', 'Gender', 'sex', 'Sex', 'SEX')
  for (nm in candidates) {
    if (nm %in% names(df)) {
      vals <- as.character(df[[nm]])
      vals <- trimws(vals)
      replace_idx <- is.na(resolved) | resolved == ''
      resolved[replace_idx] <- vals[replace_idx]
    }
  }
  resolved[resolved %in% c('M','MALE','Male','male','1')] <- 'Male'
  resolved[resolved %in% c('F','FEMALE','Female','female','0')] <- 'Female'
  resolved
}

resolve_race_column <- function(df) {
  if (nrow(df) == 0) return(character())
  resolved <- if ('race' %in% names(df)) as.character(df$race) else rep(NA_character_, nrow(df))
  candidates <- c('Race', 'race_clean', 'Race_clean')
  for (nm in candidates) {
    if (nm %in% names(df)) {
      vals <- as.character(df[[nm]])
      vals <- trimws(vals)
      replace_idx <- is.na(resolved) | resolved == ''
      resolved[replace_idx] <- vals[replace_idx]
    }
  }
  resolved[resolved %in% c('', 'NA')] <- NA_character_
  resolved
}

collapse_race_group <- function(values) {
  lowered <- tolower(trimws(as.character(values)))
  collapsed <- rep(NA_character_, length(lowered))

  collapsed[lowered %in% c('white')] <- 'White'
  collapsed[lowered %in% c('black')] <- 'Black'
  collapsed[lowered %in% c('hispanic', 'latina', 'latinx', 'mexican')] <- 'Hispanic/Latinx'
  collapsed[lowered %in% c('asian', 'filipino', 'vietnamese')] <- 'Asian'
  collapsed[lowered %in% c('multiracial', 'native american', 'native indian', 'persian')] <- 'Other/Indigenous/Mixed'

  missing_idx <- lowered %in% c('', 'na', 'nan', 'unknown', 'n/a')
  collapsed[missing_idx] <- NA_character_

  # Any leftover non-empty values roll into the grouped "other" bucket.
  other_idx <- is.na(collapsed) & !missing_idx
  collapsed[other_idx] <- 'Other/Indigenous/Mixed'
  collapsed
}

safe_cor_frame <- function(df, x_col, y_col, label) {
  needed <- c(x_col, y_col)
  if (!all(needed %in% names(df))) {
    return(tibble(metric = label, n = 0L, correlation = NA_real_))
  }
  tmp <- df %>% select(all_of(needed)) %>% drop_na()
  tibble(
    metric = label,
    n = nrow(tmp),
    correlation = if (nrow(tmp) < 2) NA_real_ else cor(tmp[[x_col]], tmp[[y_col]], use = 'complete.obs')
  )
}

safe_lm_summary <- function(df, response, predictors, label) {
  needed <- c(response, predictors)
  if (!all(needed %in% names(df))) {
    return(tibble(model = label, n = 0L, r_squared = NA_real_, adj_r_squared = NA_real_,
                  term = NA_character_, estimate = NA_real_, std.error = NA_real_, p.value = NA_real_))
  }
  fit_df <- df %>% select(all_of(needed)) %>% drop_na()
  if (nrow(fit_df) < max(4, length(predictors) + 2)) {
    return(tibble(model = label, n = nrow(fit_df), r_squared = NA_real_, adj_r_squared = NA_real_,
                  term = NA_character_, estimate = NA_real_, std.error = NA_real_, p.value = NA_real_))
  }
  formula <- stats::reformulate(predictors, response = response)
  fit <- lm(formula, data = fit_df)
  broom::tidy(fit) %>%
    mutate(
      model = label,
      n = nrow(fit_df),
      r_squared = summary(fit)$r.squared,
      adj_r_squared = summary(fit)$adj.r.squared
    ) %>%
    select(model, n, r_squared, adj_r_squared, term, estimate, std.error, p.value)
}
```

# Section 03 — Loading Dataset

```{r load-data}
override_path <- Sys.getenv("THUNDERBOLTS_MAIN", unset = "")
default_paths <- c(
  'MRI_data/analysis/tests/v12_kwsn/SN_nmdata_irredeemable_v12_kwsn_KW_legacy_clamp_prob15_kwsn_KW_legacy_clamp_merged_with_covariates.txt',
  'MRI_data/analysis/tests/v9/KW/ptfix/clamp/SN_nmdata_irredeemable_v9_KW_ptfix_clamp_prob15_KW_ptfix_clamp_with_covariates.txt'
)
metrics_path <- if (nzchar(override_path) && file.exists(override_path)) {
  override_path
} else {
  resolved <- first_existing(default_paths)
  if (is.na(resolved)) stop('No Thunderbolts input table found; set THUNDERBOLTS_MAIN.')
  resolved
}
message(glue('[Thunderbolts_v5_CPonly] Loading {metrics_path}'))
metrics <- read_delim_auto(metrics_path) %>% mutate(sub = as.character(sub))

numeric_cols <- intersect(
  c(
    'CNR_CP_corrected', 'CNR_CP_raw',
    'intensity_corrected', 'intensity_raw',
    'intensity_maximum', 'intensity_minimum',
    'age', 'years_education', 'MOCA', 'moca', 'MOCA_memory',
    'trailsb', 'nart_raw', 'NART_raw', 'SBT'
  ),
  names(metrics)
)
if (length(numeric_cols) > 0) {
  metrics <- metrics %>% mutate(across(all_of(numeric_cols), as.numeric))
}

metrics <- metrics %>%
  mutate(
    gender_clean = clean_factor(resolve_gender_column(.)),
    race_clean = clean_factor(resolve_race_column(.))
  ) %>%
  mutate(
    race_group_collapsed = clean_factor(collapse_race_group(as.character(race_clean))),
    education_years = resolve_numeric_column(., c('years_education')),
    moca_score = resolve_numeric_column(., c('MOCA', 'moca')),
    trailsb_score = resolve_numeric_column(., c('trailsb')),
    nart_score = resolve_numeric_column(., c('NART_raw', 'nart_raw'))
  )

age_breaks <- c(18, 30, 50, 60, 70, Inf)
age_labels <- c('18-30','30-50','50-60','60-70','70+')
metrics <- metrics %>% mutate(
  age_group = cut(age, breaks = age_breaks, include.lowest = TRUE, right = FALSE, labels = age_labels)
)
```

```{r dataset-summary}
tibble(
  rows = nrow(metrics),
  unique_subjects = dplyr::n_distinct(metrics$sub),
  cp_corrected_n = sum(!is.na(metrics$CNR_CP_corrected) & !is.na(metrics$intensity_corrected)),
  cp_raw_n = if ('CNR_CP_raw' %in% names(metrics) && 'intensity_raw' %in% names(metrics)) {
    sum(!is.na(metrics$CNR_CP_raw) & !is.na(metrics$intensity_raw))
  } else {
    0L
  }
)
```

```{r covariate-coverage}
tibble(
  age_n = sum(!is.na(metrics$age)),
  gender_n = sum(!is.na(metrics$gender_clean)),
  race_raw_n = sum(!is.na(metrics$race_clean)),
  race_collapsed_n = sum(!is.na(metrics$race_group_collapsed)),
  education_n = sum(!is.na(metrics$education_years)),
  moca_n = sum(!is.na(metrics$moca_score)),
  trailsb_n = sum(!is.na(metrics$trailsb_score)),
  nart_n = sum(!is.na(metrics$nart_score))
)
```

```{r race-collapse-audit}
metrics %>%
  count(race_clean, race_group_collapsed, sort = TRUE)
```

# Section 04 — CP-Only CNR and Intensity

## 04.1 Corrected CP CNR vs corrected SN intensity

```{r cp-cnr-intensity-corrected}
df_cp <- metrics %>% select(sub, intensity_corrected, CNR_CP_corrected) %>% drop_na()
ggplot(df_cp, aes(x = intensity_corrected, y = CNR_CP_corrected)) +
  geom_point(alpha = 0.6, colour = '#0072B2') +
  geom_smooth(method = 'lm', se = FALSE, colour = '#D55E00') +
  labs(
    title = 'CP CNR vs corrected SN intensity',
    x = 'SN intensity (corrected)',
    y = 'CNR (SN-control / CP)'
  ) +
  gg_theme_sn_asr
```

```{r describe-cp-cnr-intensity-corrected, results='asis'}
describe_plot(
  'CP CNR vs corrected SN intensity',
  'Each point is one subject. The red line is the best-fitting linear trend between corrected SN intensity and the CP-based CNR estimate.'
)
```

## 04.2 Raw CP CNR vs raw SN intensity

```{r cp-cnr-intensity-raw}
if (!all(c('intensity_raw', 'CNR_CP_raw') %in% names(metrics))) {
  cat('Raw CP columns unavailable; skipping raw CP scatter.\n')
} else {
  df_cp_raw <- metrics %>% select(sub, intensity_raw, CNR_CP_raw) %>% drop_na()
  ggplot(df_cp_raw, aes(x = intensity_raw, y = CNR_CP_raw)) +
    geom_point(alpha = 0.6, colour = '#7F7F7F') +
    geom_smooth(method = 'lm', se = FALSE, colour = '#E69F00') +
    labs(
      title = 'CP CNR vs raw SN intensity',
      x = 'SN intensity (raw zeropad)',
      y = 'CNR (SN-control / CP, raw)'
    ) +
    gg_theme_sn_asr
}
```

## 04.3 Correlation summary

```{r cp-correlation-summary}
bind_rows(
  safe_cor_frame(metrics, 'intensity_corrected', 'CNR_CP_corrected', 'Corrected intensity vs corrected CP CNR'),
  safe_cor_frame(metrics, 'intensity_raw', 'CNR_CP_raw', 'Raw intensity vs raw CP CNR'),
  safe_cor_frame(metrics, 'intensity_maximum', 'CNR_CP_corrected', 'Corrected intensity maximum vs corrected CP CNR'),
  safe_cor_frame(metrics, 'intensity_minimum', 'CNR_CP_corrected', 'Corrected intensity minimum vs corrected CP CNR')
)
```

```{r describe-cp-correlation-summary, results='asis'}
describe_test(
  'CP-only CNR vs intensity coupling',
  'Quantify how tightly SN intensity tracks the CP-based CNR measurement once PT, Brainstem, and nigrosome analyses are removed.',
  'Pearson correlations are reported for corrected and raw CP-specific CNR, plus the maximum and minimum corrected SN intensity summaries.'
)
```

## 04.4 Negative control

```{r cp-negative-control}
set.seed(20260227)
neg_metrics <- metrics %>% mutate(intensity_nc = sample(intensity_corrected))
safe_cor_frame(neg_metrics, 'intensity_nc', 'CNR_CP_corrected', 'Shuffled corrected intensity vs corrected CP CNR')
```

```{r describe-cp-negative-control, results='asis'}
describe_test(
  'CP-only negative control',
  'Check that the corrected CP relationship disappears when corrected SN intensity is shuffled across subjects.',
  'The CP CNR column is left untouched while the corrected SN intensity vector is permuted.'
)
```

# Section 05 — CP Regression Diagnostics

```{r cp-regression-diagnostics}
cp_models <- bind_rows(
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('intensity_corrected'), 'CP corrected ~ corrected intensity'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('intensity_corrected', 'age'), 'CP corrected ~ corrected intensity + age'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('intensity_corrected', 'age', 'gender_clean', 'race_group_collapsed'), 'CP corrected ~ corrected intensity + age + gender + collapsed race')
)
cp_models
```

```{r describe-cp-regression-diagnostics, results='asis'}
describe_test(
  'CP-only regression diagnostics',
  'Measure how much variance in CP CNR is explained by corrected SN intensity alone and after basic demographics are added.',
  'Linear models are fit only when the required columns are present with enough complete rows.'
)
```

# Section 06 — Age Analyses

## 06.1 Age vs corrected SN intensity

```{r age-vs-intensity}
age_intensity_df <- metrics %>% select(sub, age, intensity_corrected) %>% drop_na()
ggplot(age_intensity_df, aes(x = age, y = intensity_corrected)) +
  geom_point(alpha = 0.6, colour = '#0072B2') +
  geom_smooth(method = 'lm', se = FALSE, colour = '#D55E00') +
  labs(title = 'Age vs corrected SN intensity', x = 'Age', y = 'Corrected SN intensity') +
  gg_theme_sn_asr
```

## 06.2 Age vs corrected CP CNR

```{r age-vs-cp}
age_cp_df <- metrics %>% select(sub, age, CNR_CP_corrected) %>% drop_na()
ggplot(age_cp_df, aes(x = age, y = CNR_CP_corrected)) +
  geom_point(alpha = 0.6, colour = '#0072B2') +
  geom_smooth(method = 'lm', se = FALSE, colour = '#D55E00') +
  labs(title = 'Age vs corrected CP CNR', x = 'Age', y = 'Corrected CP CNR') +
  gg_theme_sn_asr
```

```{r age-models}
bind_rows(
  safe_lm_summary(metrics, 'intensity_corrected', c('age'), 'Corrected SN intensity ~ age'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age'), 'Corrected CP CNR ~ age'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed'), 'Corrected CP CNR ~ age + gender + collapsed race')
)
```

# Section 07 — Demographic Stratification

## 07.1 CP CNR by gender

```{r cp-by-gender}
gender_df <- metrics %>% select(sub, gender_clean, CNR_CP_corrected) %>% drop_na()
if (nrow(gender_df) == 0) {
  cat('Gender or CP CNR unavailable; skipping gender panel.\n')
} else {
  ggplot(gender_df, aes(x = gender_clean, y = CNR_CP_corrected, fill = gender_clean)) +
    geom_boxplot(alpha = 0.4, outlier.shape = NA) +
    geom_jitter(width = 0.12, alpha = 0.5, size = 1) +
    labs(title = 'Corrected CP CNR by gender', x = 'Gender', y = 'Corrected CP CNR') +
    gg_theme_sn_asr
}
```

## 07.2 CP CNR by collapsed race

```{r cp-by-race}
race_df <- metrics %>% select(sub, race_group_collapsed, CNR_CP_corrected) %>% drop_na()
if (nrow(race_df) == 0) {
  cat('Race or CP CNR unavailable; skipping race panel.\n')
} else {
  ggplot(race_df, aes(x = race_group_collapsed, y = CNR_CP_corrected, fill = race_group_collapsed)) +
    geom_boxplot(alpha = 0.4, outlier.shape = NA) +
    geom_jitter(width = 0.12, alpha = 0.5, size = 1) +
    labs(title = 'Corrected CP CNR by collapsed race', x = 'Collapsed race', y = 'Corrected CP CNR') +
    theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    gg_theme_sn_asr
}
```

```{r demographic-models}
bind_rows(
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('gender_clean'), 'Corrected CP CNR ~ gender'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('race_group_collapsed'), 'Corrected CP CNR ~ collapsed race'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('gender_clean', 'age'), 'Corrected CP CNR ~ gender + age'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('race_group_collapsed', 'age'), 'Corrected CP CNR ~ collapsed race + age'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed'), 'Corrected CP CNR ~ age + gender + collapsed race')
)
```

```{r describe-demographic-models, results='asis'}
describe_test(
  'Collapsed-race demographic contrasts',
  'Re-test the CP demographic contrasts after collapsing the sparse free-text race labels into coarser groups with usable sample sizes.',
  'The grouped race labels are derived from the raw race column because the NIH race/ethnicity fields are blank in this merged v12 table.'
)
```

# Section 08 — Cognition Models For CP CNR

```{r cp-cognition-models}
bind_rows(
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'education_years'), 'Corrected CP CNR ~ age + gender + collapsed race + education'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'moca_score'), 'Corrected CP CNR ~ age + gender + collapsed race + MOCA'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'trailsb_score'), 'Corrected CP CNR ~ age + gender + collapsed race + Trails B'),
  safe_lm_summary(metrics, 'CNR_CP_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'nart_score'), 'Corrected CP CNR ~ age + gender + collapsed race + NART')
)
```

```{r describe-cp-cognition-models, results='asis'}
describe_test(
  'CP cognition-adjusted models',
  'Ask whether education or cognition shifts the corrected CP CNR signal after the core demographic terms are already in the model.',
  'Each cognition variable is added one at a time on top of age, gender, and the collapsed race grouping.'
)
```

# Section 09 — Corrected Intensity Models

## 09.1 Intensity demographics

```{r intensity-demographic-models}
bind_rows(
  safe_lm_summary(metrics, 'intensity_corrected', c('age'), 'Corrected SN intensity ~ age'),
  safe_lm_summary(metrics, 'intensity_corrected', c('gender_clean'), 'Corrected SN intensity ~ gender'),
  safe_lm_summary(metrics, 'intensity_corrected', c('race_group_collapsed'), 'Corrected SN intensity ~ collapsed race'),
  safe_lm_summary(metrics, 'intensity_corrected', c('age', 'gender_clean', 'race_group_collapsed'), 'Corrected SN intensity ~ age + gender + collapsed race')
)
```

## 09.2 Intensity cognition-adjusted models

```{r intensity-cognition-models}
bind_rows(
  safe_lm_summary(metrics, 'intensity_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'education_years'), 'Corrected SN intensity ~ age + gender + collapsed race + education'),
  safe_lm_summary(metrics, 'intensity_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'moca_score'), 'Corrected SN intensity ~ age + gender + collapsed race + MOCA'),
  safe_lm_summary(metrics, 'intensity_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'trailsb_score'), 'Corrected SN intensity ~ age + gender + collapsed race + Trails B'),
  safe_lm_summary(metrics, 'intensity_corrected', c('age', 'gender_clean', 'race_group_collapsed', 'nart_score'), 'Corrected SN intensity ~ age + gender + collapsed race + NART')
)
```

```{r describe-intensity-models, results='asis'}
describe_test(
  'Corrected intensity demographic and cognition models',
  'Mirror the CP models using corrected SN intensity itself as the outcome so we can tell whether any demographic or cognition effects are intensity-driven rather than CNR-driven.',
  'The same collapsed race grouping and one-at-a-time cognition additions are reused for the intensity models.'
)
```

# Section 10 — Session Notes

This CP-only Thunderbolts variant intentionally excludes PT-control, Brainstem,
and all nigrosome analyses. It is meant for a focused readout of the
SN-control/CP background relationship on the supplied Irredeemable table.
